ProteInfer: deep networks for predicting functional properties of proteins

### 2021-09-20 latex/the_supplement.tex
Updates

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for predicting functional properties of proteins}
\shorttitle{ProteInfer}

\author[1,\Letter]{Authors}
\affil[1]{Google}

\maketitle

\begin{abstract}
Predicting the function(s) of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. In our work, we instead employ deep convolutional neural networks to directly predict a variety of protein functions - EC numbers and GO terms - directly from an unaligned amino acid sequence. This performs comparably to alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces. In response, we have provided an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract
\begin{keywords}
neural networks | protein | prediction | classification | embedding
\end{keywords}
\begin{corrauthor}
\todo{x} \at google.com
\end{corrauthor}
\section*{Introduction}
Every day, more than a hundred thousand protein sequences are added to global sequence databases \cite{tremblstats}. However, these entries are of limited use to practitioners unless they are accompanied by functional annotations. While curators diligently extract annotations from the literature, assessing more than 60,000 papers each year \cite{uniprotlatest}, the time-consuming nature of this task means that only 0.03\% of publicly available protein sequences are manually annotated.
%%%FIGURETREMBL%%%
The community has a long history of using computational tools to infer protein function directly from amino acid sequence. Starting in the 1980s, methods such as BLAST \cite{blast} relied on pairwise sequence comparisons, where a query protein is assumed to have the same function as highly similar sequences that have already been annotated. Signature-based approaches were later introduced, with the PROSITE database \cite{bairoch1991prosite} cataloguing short amino acid ``motifs'' found in proteins that share a particular function. Subsequently, a crucial refinement of signature-based approaches was the development of profile hidden Markov models (HMMs) \cite{krogh1994hidden, eddyhmmreview}. These models collapse an alignment of related protein sequences into a model that provides likelihood scores for new sequences, that describe how well they fit the aligned set. Critically, profile HMMs allow for longer signatures and fuzzier matching and are currently used to update popular databases such as Interpro and Pfam \cite{pfamlatest, blum2021interpro}. Subsequent refinements have made these techniques more sensitive and computationally efficient \cite{altschul1997gapped, fasterhmmer, pfamlatest, soding2005protein, steinegger2019hh}, while their availability as web tools allows practitioners to easily incorporate them into workflows \cite{johnson2008ncbi, soding2005hhpred, potter2018hmmer, bernhofer2021predictprotein}.
\begin{figure}[t]
  \includegraphics[width=\columnwidth]{Figures/new_the_task.pdf}%
  \caption{\textbf{Predicting functional properties from amino acid sequence has traditionally been achieved by comparing to many sequences (BLAST) or many models (InterProScan), deep learning models provide an alternative approach.}  A single neural network is trained to predict a large number of functional properties from an amino acid sequence.\label{fig:task}}
These computational modelling approaches have had great impact; however, one third of bacterial proteins still cannot be annotated (even computationally) with a function \cite{phenotypes}. It is therefore worthwhile to examine how new approaches might complement existing techniques. First, current approaches conduct entirely separate comparisons for each comparator sequence or model, and thus may not fully exploit the features shared across different functional classes. An ideal classification system, for example, might have a modular ATP-binding region detector used in detection of both kinases and ABC transporters \citep{pmid12468712}. Separately modelling these targets in each family, like standard HMM approaches, increases the computational cost and may also be less accurate. In addition, the process of creating many of these signatures is not fully automated and requires considerable curatorial efforts \citep{pfamprofilebuilding, pfamrepeats}, which at present are spread across an array of disparate but overlapping signature databases \citep{blum2021interpro}.

%%%FIGURESVG%%%

By analogy to deep networks for images, the early layers in these models build up an understanding of simple features such as edges, and later layers use these features to identify textures, and then entire objects. Edge detecting filters can thus be trained with information from all the labelled examples, and the same filters can be used to detect, for instance, both oranges and lemons \cite{activationatlas}.

In response, recent work has contributed a number of deep neural network models for protein function classification\citep{ecpred,deepgo,prolango,deeploc,dspace,deepred,deepre,deepsf,littmann2021embeddings}. These approaches train a single model to recognise multiple properties, building representations of different protein functions via a series of layers, which allow the same low-level features to be used for different high-level classifications. Of special note is the layer preceding the final layer of the network, which constructs an ``embedding'' of the entire example in a high-dimensional vector space, and often captures semantic features of the input.

Beyond functional annotation, deep learning has enabled significant advances in protein structure prediction \cite{alquraishi2019end, senior2020improved, Yang1496, yang2020improved, du2020energy, rao2021msa}, predicting the functional effects of mutations \cite{riesselman2018deep,rives2019biological,rao2019evaluating,frazer2020large}, and protein design \cite{yang2019machine, mazurenko2019machine, biswas2020low, madani2020progen,elnaggar2020prottrans, anishchenko2020novo, bryant2021deep}. A key departure from traditional approaches is that researchers have started to incorporate vast amounts of raw, uncurated sequence data into model training.

Of particular relevance to the present work is Bileschi et al. (2019) \citep{bileschi}, where it is shown that models with residual layers \citep{resnet} of dilated convolutions \citep{dilatedconvolutions} can categorise protein domains more accurately than profile HMMs or BLAST. Dohan (2021) \cite{dohan2021kdd} provides additional accuracy improvements using uncurated data. However, these models cannot infer functional annotations for full-length protein sequences, since they are trained on pre-segmented domains and can only predict a single label. The full-sequence task is of primary importance to biological practitioners.

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Methods}
\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{Figures/figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold; varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the Supplement.
\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are further methods for splitting, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. A drawback of this approach is that it is not available for EC numbers, test sets are often small and have potential biases (e.g., they are limited to a small number of species or functions). Further, there is no such dataset for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives\citep{warwick2020benchmarking}.

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}
Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in Supplement). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

BLASTp is arguably a practitioner's default choice for functional annotation, so we implemented an alignment-based baseline in which BLASTp is used to identify the closest sequence to a query sequence in the train set. Labels are then imputed for the query sequence by transferring those labels that apply to the annotated match from the train set.

We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score. %
%startleaveout
(See \Cref{sup:fig:prc_ec_random}, \Cref{sup:fig:prc_go_random},
\Cref{sup:fig:prc_ec_clustered}, \Cref{sup:fig:prc_go_clustered}.)
%stopleaveout

%%%METHOD%%%
We found that BLASTp was able to achieve higher recall values than ProteInfer for lower precision values, while ProteInfer was able to provide greater precision than BLASTp at lower recall values. We wondered whether a combination of ProteInfer and BLASTp could synergize the best properties of both approaches. We found that even the simple ensembling strategy of rescaling the BLAST bit score by the averages of the ensembled CNNs' predicted probabilities gave a $\texttt{F}_{max}$ score (0.991, 95\% confidence interval [CI]: 0.990--0.992 ) that exceeded that of BLAST (0.984, 95\% CI: 0.983--0.985) or the ensembled CNN (0.981, 95\% CI: 0.980--0.982) alone. 
On the clustered train-test split based on UniRef50 (see \textit{clustered} in \Cref{fig:toplevel}B%htmlver:<a href="#fig5" target="_self">Fig. 5</a>
), we see a performance drop in all methods: this is expected, as remote homology tasks are designed to challenge methods to generalize farther in sequence space. The $\texttt{F}_{max}$ score of a single neural network fell to 0.914 (95\% CI: 0.913--0.915, precision: 0.959 recall: 0.875), substantially lower than BLAST (0.950, 95\% CI: 0.950--0.951), though again an ensemble of both BLAST and ProteInfer outperformed both (0.979, 95\% CI: 0.979--0.980). We find that neural network methods learn different information about proteins than alignment-based methods, and a combination of the two further improves remote homology detection.

We also examined the relationship between the number of examples of a label in the training dataset and the performance of the model. In an image recognition task, this is an important consideration since one image of, say, a dog, can be utterly different to another. Large numbers of labels are therefore required to learn filters that are able to predict members of a class. In contrast, for sequence data we found that even for labels that occurred less than five times in the training set, 58\% of examples in the test set were correctly recalled, while achieving a precision of 88\%, for an F1 of 0.7 (\Cref{trainset}%htmlver: <a href="#fig6" target="_self">Fig. 6</a>
). High levels of performance are maintained with few training examples because of the evolutionary relationship between sequences, which means that one ortholog of a gene may be similar in sequence to another (\Cref{seq_similarity}%htmlver: 
). The simple BLAST implementation described above also performs well, and better than a single neural network, likely again exploiting the fact that many sequence have close neighbours in sequence space with similar functions. We again find that ensembling the BLAST and ProteInfer outputs provides performance exceeding that of either technique used alone.
%%%NUM_TRAIN_EGS%%%
\subsection*{Deep models link sequence regions to function}
%%%cam_colab%%% %
Proteins that use separate domains to carry out more than one enzymatic function are particularly useful in interpreting the behaviour of our model. For example, \textit{S. cerevisiae} fol1 (accession Q4LB35) catalyses three sequential steps of tetrahydrofolate synthesis, using three different protein domains (\Cref{fig:cam}A%htmlver:
). This protein is in our held-out test set, so no information about its labels was directly provided to the model.

To investigate what sequence regions the neural network is using to make its functional predictions, we used class activation mapping (CAM) \citep{classactivationmapping} to identify the sub-sequences responsible for the model predictions. We found that separate regions of sequence cause the prediction of each enzymatic activity, and that these regions correspond to the known functions of these regions (\Cref{fig:cam}B%htmlver: <a href="#fig7" target="_self">Fig. 7</a>
).  This demonstrates that our network identifies relevant elements of a sequence in determining function.

%%%3FUNCTIONB%%%

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering (see Supplement for details of the method)%htmlver:see Supplement
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity (see Supplement)%htmlver:see Supplement 
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

To visualise this space, we selected proteins with a single leaf-level EC number and used UMAP to compress their embeddings into two dimensions \citep{umap}.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{Figures/ec_embeddings.png}%
  \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
  \label{fig:umap}
\end{figure*}
%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} We took the embeddings for all non-enzymatic sequences in the test set and projected them using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, Q8RUD6 is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). Q8RUD6 can however act as a sulfertransferase \cite{pmid17408957}.

Note that the model is directly trained with labels reflecting the EC hierarchy; the structure in \Cref{fig:umap} %htmlver: <a href="#fig8" target="_self">Fig. 8</a>
 was not discovered automatically from the data. However, we can also ask whether the embedding captures more general protein characteristics, beyond those on which it was directly supervised.

To investigate this, we took the subset of proteins in Swiss-Prot that are non-enzymes, and so lack any EC annotations. The network would achieve perfect accuracy on these examples if it e.g. mapped all of them to a single embedding that corresponds to zero predicted probability for every enzymatic label. Do these proteins therefore share the same representation in embedding space? The UMAP projection of these sequences' embeddings revealed clear structure to the embedding space, which we visualised by highlighting several GO annotations which the network was never supervised on. For example, one region of the embedding space contained ribosomal proteins, while other regions could be identified containing nucleotide binding proteins, or membrane proteins (\Cref{fig:unsupervised_embedding}%htmlver:interactive figure below
).
%%%nonenzymes%%%
To quantitatively measure whether these embeddings capture the function of non-enzyme proteins, we trained a simple random forest classification model to predict whether a protein was annotated with the \textit{intrinsic component of membrane} GO term. We trained on 100 non-enzymes as above\footnote{18.6\% of the non-enzymes were membrane proteins, and so when repeating this procedure with different splits, there were usually $\leq 20$ positive examples among the 100}, evaluated on the rest of the examples, and achieved a precision of 97\% and recall of 65\% for an F1 score of 0.78. Model training and data-labelling typically took around 10 seconds for smaller models. This demonstrates the power of embeddings to simplify other studies with limited labeled data, as has been observed in recent work \citep{Alley589333,biswas2020low}.
%%%SWEEP_embedding%%%

\subsection*{Rapid client-side in-browser protein function prediction}
Processing speed and ease-of-access are important considerations for the utility of biological software. An algorithm that takes hours or minutes is less useful than one that runs in seconds, both because of its increased computational cost, but also because it allows less immediate interactivity with a researcher. An ideal tool for protein function prediction would require minimal installation and would instantly answer a biologist's question about protein function, allowing them to immediately act on the basis of this knowledge. Moreover, there may be intellectual property concerns in sending sequence data to remote servers, so a tool that does annotation completely client-side may also be preferable.

There is arguably room for improvement in this regard from classical approaches. For example, the online interface to InterProScan can take ~147 seconds to process a 1500 amino acid sequence\footnote{Protein Q77Z83.}\footnote{It should be noted that the individual databases that make up InterProScan may return matches faster, with the online interface to Pfam taking 14--20 seconds for a 1500 amino acid sequence.}, while running the tool may make the search faster, doing so requires downloading a 9 GB file, with an additional 14 GB for the full set of signatures. When uncompressed this installation exceeds 51 GB. Meanwhile, conducting a BLAST search against Swiss-Prot takes 34 seconds for a 1500 amino acid sequence \footnote{A target database with decreased redundancy could be built to reduce this search time, and other optimizations of BLAST have been developed.}.

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see Supplement
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}
\begin{figure*}[b]
\centering
  \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
  % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
  \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
  \label{fig:cam}
\end{figure*}

To assess our model's performance using an additional experimentally-validated source of ground truth, we focused our attention on a large set of bacterial genes for which functions have recently been identified in a high-throughput experimental genetic study \citep{phenotypes}. In particular, this study listed newly identified EC numbers for 171 proteins, representing cases when there was previously either misannotation or inconsistent annotation in the SEED or KEGG databases. Therefore, this set of genes may be enriched for proteins whose functions are difficult to assess computationally.
We examined how well our network was able to make predictions for this experimental dataset at each level of the EC hierarchy (\Cref{fig:levels}%htmlver: <a href="#fig10" target="_self">Fig. 11</a>
), using as a decision threshold the value that optimised F1 identified during tuning. The network had high accuracy for identification of broad enzyme class, with 90\% accuracy at the top level of the EC hierarchy. To compute accuracy, we examined the subset of these 171 proteins for which there was a single enzymatic annotation from \cite{phenotypes}, giving us predictions for 119 enzymes. At the second level of the hierarchy, accuracy was 90\% and the network declined to make a prediction for 12\% of classes. Even at the third level, accuracy was 86\% with the network making a prediction in 77\% of cases. At the finest level of classification, the proportion of examples for which a prediction was made fell to 28\%, with 42\% of these predictions correct.

As an  example, the \textit{Sinorhizobium meliloti} protein Q92SI0 is annotated in UniProt as a Inosine-uridine nucleoside N-ribohydrolase (EC 3.2.2.1).  Analysing the gene with InterProScan \citep{interproscan} also gives this prediction, but our model instead predicts it to be a uridine nucleosidase (EC 3.2.2.3), and this was indeed the result found in this experimental work. Similarly, \textit{Pseudomonas fluorescens} A0A166Q345 was correctly classified by our model as a D-galacturonate dehydrogenase (EC 1.1.1.203) as opposed to a misannotation on UniProt and with InterProScan.

%%%PRICE%%%
It was notable that for many of these proteins, the network declined to make a prediction at the finest level of the EC hierarchy. This suggests that by training on this hierarchical data, the network is able to appropriately make broad or narrow classification decisions. This is similar to the procedure employed with manual annotation: when annotators are confident of the general class of reaction that an enzyme catalyses but not its specific substrate, they may leave the third or fourth position of the EC number blank (e.g. EC:1.1.-.-). Due to training on hierarchical data, our network is able to reproduce these effects by being more confident (with higher accuracy) at earlier levels of classification.

\subsection*{A model predicting the entire gene ontology}
Given the high accuracy that our deep learning model was able to achieve on the more than five thousand enzymatic labels in Swiss-Prot, we asked whether our networks could learn to predict protein properties using an even larger vocabulary of labels, using a similar test-train setup. Gene Ontology \citep{ashburner2000gene, gene2019gene, carbon2009amigo} (GO) terms describe important protein functional properties, with 28,079 such terms in Swiss-Prot that cover the molecular functions of proteins (e.g. DNA-binding, amylase activity), the biological processes they are involved in (e.g. DNA replication, meiosis), and the cellular components to which they localise (e.g. mitochondrion, cytosol). These terms are arranged in a complex directed acyclic graph, with some nodes having as many as 12 ancestral nodes.

We note that there has been extensive work in GO label prediction evaluated on a temporally-split dataset (constructing a test set with the most recently experimentally annotated proteins), e.g., \citep{zhou2019cafa}, and stress that our comparison is instead on random and clustered splits of Swiss-Prot.  As discussed above, our approaches to splitting the data into train and test has different advantages and disadvantages as compared to a temporal split.
%%%GO_METHOD%%%
We trained a single model to predict presence or absence for each of these terms and found that our network was able to achieve a precision of 0.918 and a recall of 0.854 for an F1 score of 0.885 (95\% CI: 0.882--0.887).

An ensemble of multiple CNN elements was again able to achieve a slightly better result with an F1 score of 0.899 (95\% CI: 0.897--0.901), which was exceeded by a simple transfer of the BLAST top pick at 0.902 (95\% CI: 0.900--0.904), with an ensemble of both producing the best result of 0.908 (95\% CI: 0.906--0.911).

To benchmark against a common signature-based methodology, we used InterProScan to assign protein family signatures to each test sequence. We chose InterProScan for its coverage of labels as well as its use of multiple profile-based annotation methods, including HMMER and PROSITE, mentioned above. We note that while InterProScan predicts GO labels directly, it does not do so for EC labels, which is why we did not use InterProScan to benchmark our work on predicting EC labels.

We found that InterProScan was largely fairly precise, but did not recall many labels, giving it a precision of 0.937 and recall of 0.543 for an F1 score of 0.688. We note that both ProteInfer's recall at a precision of .937 is substantially higher (0.835) than InterProScan at assigning GO labels.

>>>>

\section*{Supplement}
\beginsupplement

\subsection{Implementation details}

\subsubsection*{Label inheritance}
Some annotations, such as GO terms and EC numbers, come as directed acyclic graphics (which take the form simple trees for EC numbers). Typically in such cases the annotation provided on UniProt is the most-specific that is known. For example if a protein is known to exhibit \textit{sequence specific DNA binding} (GO:0043565). It will not separately be annotated with the ancestral term \textit{DNA binding} (GO:0003677), this is simply assumed from the ontology. Using such an annotations directly, however, is likely to be problematic in a deep learning setting. Failing to annotate an example with the parental term demands that the model predict that the example is negative for this term, which is not the effect we want.
To address this our datasets include labels for all ancestors of applied labels for EC, GO and InterPro datasets. In the case of GO we restrict these to \textit{is\_a} relationships.

\subsubsection*{Class activation mapping}
Many proteins have multiple functional properties. For example we analyse the case of the bifunctional dhfr/ts of \textit{T. gondii}. Such bifunctional enzymes are often not unique -- it is functionally advantageous for these enzymes to be fused, which facilitates channeling of substrate between their active sites. Since there are a number of such examples within Swiss-Prot, the mere existence of a TS domain in a protein is (mild) evidence for possible DHFR function. To increase the interpretability of the network we introduce a technique that simplifies such cases by subtracting the activations for other predicted classes during class activation mapping.

\subsection*{Model architecture}
To create an architecture capable of receiving a wide range of input sequences, with computational requirements determined for each inference by the length of the individual input sequence, we employed a dilated convolutional approach \citep{dilatedconvolutions}. Computation for both training and prediction in such a model can be parallelized across the length of the sequence. By training on full length proteins, in a multi-label training setting, we aimed to build networks that could  extract functional information from raw amino acid sequences. One helpful feature of this architecture is its flexibility with regards to sequence length. Natural protein sequences can vary in length by at least three orders of magnitude, but some architectures have computational requirements that scale with the maximum sequence they are capable of receiving as input, rather than the sequence being currently examined. These fixed-length approaches reduce efficiency as well as place a hard limit on the length of sequences that can be examined.
